\begin{abstract}
Multi-step workflows that chain large language model (LLM) calls suffer from \textit{context pollution}: as information accumulates across steps, models hallucinate, confuse intermediate outputs, and lose track of task constraints. We present \textbf{NormCode}, a semi-formal language for constructing \textit{plans of inferences}---structured decompositions where each step operates in data isolation, receiving only explicitly passed inputs. This eliminates cross-step contamination by construction. NormCode enforces a strict separation between \textit{semantic operations} (LLM-driven reasoning, non-deterministic) and \textit{syntactic operations} (deterministic data restructuring), enabling precise cost and reliability tracing. The language exists in three isomorphic formats: \texttt{.ncds} for human authoring, \texttt{.ncd} for machine execution, and \texttt{.ncn} for human verification---supporting progressive formalization from sketch to production. We validate NormCode through two demonstrations: (1) a base-X addition algorithm achieving 100\% accuracy on arbitrary-length inputs, and (2) self-hosted execution of NormCode's own five-phase compiler pipeline. The working orchestrator provides dependency-driven scheduling, SQLite-backed checkpointing, and loop management. By making AI workflows auditable by construction, NormCode addresses a critical need for transparency in high-stakes domains such as legal reasoning, medical decision-making, and financial analysis.
\end{abstract}
